---
title: "Fig 2"
author: "SK"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    highlight: kate
editor_options: 
  markdown: 
    wrap: 65
---


```{r, include=FALSE}
library(ggplot2)
library(ggforce)
library(ggrepel)
library(patchwork)
library(rstan)
library(parallel)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
rstan_options(threads_per_chain = 1)

knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

# Figure: CDR3 length in PT and SLN samples

```{r}
tnp <- get(load("data/TCRb_nonprod.RData"))
t <- get(load("data/TCRb.RData"))
```


First, we will look at the average CDR3 length in PT and SLN
samples.

```{r}
get_cdr3_len <- function(name, prod, t) {
  
  # PT
  q_pt <- t[t$name == name & t$tissue == "PT" & t$prod == prod, ]
  y_pt <- rep(q_pt$cdr3_length, times = q_pt$templates)
  
  # SLN
  q_sln <- t[t$name == name & t$tissue == "SLN" & t$prod == prod, ]
  y_sln <- rep(q_sln$cdr3_length, times = q_sln$templates)
  
  o <- data.frame(name = name, 
                  productive = prod,
                  mean_cdr3_pt = mean(q_pt$cdr3_length),
                  mean_cdr3_sln = mean(q_sln$cdr3_length),
                  delta_cdr3 = mean(q_pt$cdr3_length)-mean(q_sln$cdr3_length), 
                  wmean_cdr3_pt = mean(y_pt),
                  wmean_cdr3_sln = mean(y_sln),
                  wdelta_cdr3 =  mean(y_pt)-mean(y_sln),
                  days_surgery = q_pt$days_surgery[1],
                  timing = q_pt$timing[1],
                  stadium = q_pt$Stadium[1])
  return(o)
}
```

```{r}
cdr3_len <- rbind(do.call(rbind, lapply(X = unique(t$name), FUN = get_cdr3_len, t = t, prod = T)),
                  do.call(rbind, lapply(X = unique(tnp$name), FUN = get_cdr3_len, t = tnp, prod = F)))
```


```{r}
top_cdr3s <- c()
for(s in unique(t$sample_name)) {
  q <- t[which(t$sample_name == s), ]
  q <- q[order(q$templates, decreasing = T), ]
  q <- q[1:100, ]
  top_cdr3s <- rbind(top_cdr3s, q)
}

cdr3_len_top <- do.call(rbind, lapply(X = unique(top_cdr3s$name), 
                                   FUN = get_cdr3_len, t = top_cdr3s, prod = T))
cdr3_len_top$productive <- "top-100 clones"

cdr3_len <- rbind(cdr3_len, cdr3_len_top)
rm(q, s)
```


```{r, fig.width=6, fig.height=3.2, warning=FALSE}
cdr3_len$productive_label <- "top-100 clones"
cdr3_len$productive_label[cdr3_len$productive=="TRUE"] <- "productive"
cdr3_len$productive_label[cdr3_len$productive=="FALSE"] <- "non-productive"
cdr3_len$productive_label <- factor(x = cdr3_len$productive_label,
                                    levels = c("productive",
                                               "top-100 clones",
                                               "non-productive"))

g <- ggplot(data = cdr3_len)+
  facet_wrap(facets = ~productive_label)+
  geom_point(aes(y = mean_cdr3_pt, x = mean_cdr3_sln), size = 2, stroke = 0, alpha = 0.35)+
  theme_bw(base_size = 10)+
  theme(legend.position = "top", strip.text.x = element_text(margin = margin(0.05,0,0.05,0, "cm")))+
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")+
  ylab("Mean CDR3 length [nt] in PT")+
  xlab("Mean CDR3 length [nt] in SLN")+
  scale_x_continuous(breaks = scales::pretty_breaks(n=4), limits = c(39, 47))+
  scale_y_continuous(breaks = scales::pretty_breaks(n=4), limits = c(39, 47))

g
```


Most of the points lie above the diagonal: on average we see longer CDR3s in 
PT vs SLN! We are talking about a potential difference in 0.25 nt $\rightarrow$ 
small effect!

How to interpret this phenomenon?

Do we see the same pattern among CDR3s of **non-productive** TCRs? The answer
is no, i.e. this phenomenon only occurs among productive (selected) TCRs and
cannot be attributed to a sample-specific bias?

At the same time, I do not see a difference in CDR3 length associated with
IM vs DEL.


## Quantiative analysis

```{r}
model_m <- rstan::stan_model(file = "model/model_cdr3.stan")
```


## Productive TCRs

```{r}
y <- cdr3_len[cdr3_len$productive_label=="productive",]
fit_prod <- sampling(object = model_m,
                     data = list(y_pt = y$mean_cdr3_pt,
                                 y_sln = y$mean_cdr3_sln,
                                 N = nrow(y)),
                     chains = 4, 
                     iter = 2000, 
                     warmup = 1000,
                     cores = 4, 
                     refresh = -1)
```




## Productive (top-100 clones) TCRs

```{r}
y <- cdr3_len[cdr3_len$productive_label=="top-100 clones",]
fit_top_prod <- sampling(object = model_m,
                     data = list(y_pt = y$mean_cdr3_pt,
                                 y_sln = y$mean_cdr3_sln,
                                 N = nrow(y)),
                     chains = 4, 
                     iter = 2000, 
                     warmup = 1000,
                     cores = 4, 
                     refresh = -1)
```


## Non-productive TCRs

```{r}
y <- cdr3_len[cdr3_len$productive=="FALSE",]
fit_nonprod <- sampling(object = model_m,
                        data = list(y_pt = y$mean_cdr3_pt,
                                    y_sln = y$mean_cdr3_sln,
                                    N = nrow(y)),
                        chains = 4, 
                        iter = 2000, 
                        warmup = 1000,
                        cores = 4, 
                     refresh = -1)
```


```{r, fig.width=3, fig.height=4}
s <- data.frame(rbind(summary(fit_prod, par = "alpha")$summary,
           summary(fit_top_prod, par = "alpha")$summary,
           summary(fit_nonprod, par = "alpha")$summary))
s$group <- c("productive", "top-100 clones", "non-productive")
s$group <- factor(x = s$group, levels = rev(s$group))
g_cdr3_difference <- ggplot(data = s)+
  geom_vline(xintercept = 0, linetype = "dashed", col = "gray")+
  geom_point(aes(y = group, x = mean))+
  geom_errorbarh(aes(y = group, xmin = X2.5., xmax = X97.5.), height = 0.1)+
  xlab(label = expression(Delta[PT-SLN]*" (nt)"))+
  ylab(label = "")
```


## INDELS

```{r, fig.width=7, fig.height=2.5}
# productive
n1 <- aggregate(n1_insertions~tissue+name+patient_id, data = t, FUN = mean)
n1$key <- "N1 insertions"
n1$y <- n1$n1_insertions
n1$n1_insertions <- NULL

n2 <- aggregate(n2_insertions~tissue+name+patient_id, data = t, FUN = mean)
n2$key <- "N2 insertions"
n2$y <- n2$n2_insertions
n2$n2_insertions <- NULL

d3 <- aggregate(d3_deletions~tissue+name+patient_id, data = t, FUN = mean)
d3$key <- "D3 deletions"
d3$y <- d3$d3_deletions
d3$d3_deletions <- NULL

d5 <- aggregate(d5_deletions~tissue+name+patient_id, data = t, FUN = mean)
d5$key <- "D5 deletions"
d5$y <- d5$d5_deletions
d5$d5_deletions <- NULL

indels <- rbind(n1, n2, d3, d5)
rm(n1, n2, d3, d5)

indels$name <- NULL
indels_pt <- indels[indels$tissue=="PT",]
indels_sln <- indels[indels$tissue!="PT",]
indels <- merge(x = indels_pt, y = indels_sln, by = c("patient_id", "key"))
indels$productive_label <- "productive"

# productive (top-100 clone)
n1 <- aggregate(n1_insertions~tissue+name+patient_id, data = top_cdr3s, FUN = mean)
n1$key <- "N1 insertions"
n1$y <- n1$n1_insertions
n1$n1_insertions <- NULL

n2 <- aggregate(n2_insertions~tissue+name+patient_id, data = top_cdr3s, FUN = mean)
n2$key <- "N2 insertions"
n2$y <- n2$n2_insertions
n2$n2_insertions <- NULL

d3 <- aggregate(d3_deletions~tissue+name+patient_id, data = top_cdr3s, FUN = mean)
d3$key <- "D3 deletions"
d3$y <- d3$d3_deletions
d3$d3_deletions <- NULL

d5 <- aggregate(d5_deletions~tissue+name+patient_id, data = top_cdr3s, FUN = mean)
d5$key <- "D5 deletions"
d5$y <- d5$d5_deletions
d5$d5_deletions <- NULL

indels_top <- rbind(n1, n2, d3, d5)
rm(n1, n2, d3, d5)

indels_top$name <- NULL
indels_pt <- indels_top[indels_top$tissue=="PT",]
indels_sln <- indels_top[indels_top$tissue!="PT",]
indels_top <- merge(x = indels_pt, y = indels_sln, by = c("patient_id", "key"))
indels_top$productive_label <- "top-100 clones"

indels <- rbind(indels, indels_top)

g_indels <- ggplot(data = indels)+
  facet_wrap(facets = ~productive_label, nrow = 1, scales = "free_y")+
  geom_hline(yintercept = 0, linetype = "dashed")+
  geom_sina(aes(x = key, y = y.x-y.y), size = 1.25, stroke = 0, alpha = 0.35)+
  ylab(label = "PT - SLN [nt]")+
  xlab(label = '')+
  theme_bw(base_size = 10)+
  theme(legend.position = "top", strip.text.x = element_text(margin = margin(0.05,0,0.05,0, "cm")))

g_indels
```


```{r, fig.width=7.4, fig.height=5}
g_cdr3_top <- (g|g_cdr3_difference)+plot_layout(widths = c(5,1))

g_cdr3 <- (g_cdr3_top/g_indels)+plot_layout(heights = c(1,1))
g_cdr3
```


```{r}
ggsave(filename = "Fig_2.pdf",
       plot = g_cdr3,
       device = "pdf",
       width = 7,
       height = 4)
```

