---
title: "Fig 1: NRADs"
author: "SK"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    highlight: kate
editor_options: 
  markdown: 
    wrap: 65
---



```{r, include=FALSE}
library(dplyr)
library(ggplot2)
library(ggforce)
library(ggrepel)
library(RADanalysis)
library(patchwork)
```

```{r, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

```{r}
t <- get(load("data/TCRb.RData"))
```


# Diversity analysis with NRADs


## Raw RADs

Even with 'raw' RADs we already see differences in diversity
between SLN and PT. Black points are CDR3s that are identical
to melanoma specific CDR3s

I expected this:

  * black points are mostly located at the heads of the RADs $\rightarrow$ this
  is not the case
  * black points are mostly located at the heads of the RADs **in PTs** 
  $\rightarrow$ this is also not the case

```{r}
ggplot()+
  geom_point(data  = t, aes(x = rank, y = templates, col = tissue), size = 0.5)+
  geom_line(data  = t, aes(x = rank, y = templates, group = sample_name, col = tissue))+
  scale_x_log10(name = "Rank")+
  scale_y_log10(name = "RAD")+
  theme_bw()+
  theme(legend.position = "top")+
  scale_color_manual(values = c("steelblue", "orange"))
```

```{r}
sort(table(t$sample_name))
```

```{r, fig.width=5, fig.height=5}
rads <- reshape2::acast(data = t, 
                        formula = sample_name~rank, 
                        value.var = "templates", 
                        fill = 0)
```

```{r, cache=T,echo=F, message=F, comment=F, warning=F, error=F}
nrads <- RADnormalization_matrix(input = rads, 
                                 max_rank = 2079,
                                 average_over = 500, 
                                 verbose = F)
```


```{r, echo=F, message=F, comment=F, warning=F, error=F}
nrad_data <- nrads$norm_matrix
rownames(nrad_data) <- rownames(rads)
nrad_data <- reshape2::melt(nrad_data)
colnames(nrad_data) <- c("rownames", "rank", "norm.abundance")
nrad_data$sample_name <- as.character(nrad_data$rownames)
nrad_data$rownames <- NULL

w <- merge(y = t[duplicated(t$sample_name)==F, c("name", "sample_name", "tissue", "timing")],
           x = nrad_data, by.y = "sample_name", by.x = "sample_name", all.x = T)
```


## Shannon entropy and diversity

```{r}
hs <- c()
for(s in unique(t$sample_name)) {
  x <- t[t$sample_name == s, ]
  x$p <- x$templates/sum(x$templates)
  hs <- rbind(hs, data.frame(h = -sum(x$p*log(x$p)), 
                             n_cell = sum(x$templates), 
                             n_clone = nrow(x),
                             sln_time = x$timing[1],
                             tissue = x$tissue[1],
                             sample_name = s,
                             name = x$name[1]))
}

hs$h_cell <- hs$h/log(hs$n_cell)
hs$h_clone <- hs$h/log(hs$n_clone)
```


```{r, fig.width=3.5, fig.height=3.3}
hss <- merge(x = hs[hs$tissue=="PT",c("name", "h")], 
             y = hs[hs$tissue=="SLN",c("name", "h")],
             by = "name")
g_H <- ggplot(data = hss)+
  geom_point(aes(x = h.x, y = h.y), size = 1.25, shape = 21, fill = "black", col = "white")+
  ylab(label = "H (SLN)")+
  xlab(label = "H (PT)")+
  theme(legend.position = "top")+
  scale_x_continuous(limits = c(6, 11))+
  scale_y_continuous(limits = c(6, 11))+
  geom_abline(slope = 1, intercept = 0, linetype = "dashed")

g_H
```


```{r, fig.width=3.5, fig.height=3.3}
ggsave(filename = "Fig_1_inset.pdf",
       plot = g_H,
       device = "pdf",
       width = 1.5,
       height = 1.5)
```




```{r, eval = F}
g_nrads <- (ggplot(data  = w)+
             geom_line(aes(x = rank, y = norm.abundance, group = sample_name, col = tissue))+
             scale_x_log10(name = "Rank")+
             scale_y_log10(name = "Normalized abundance")+
             annotation_logticks(base = 10, sides = c("lb"))+
             theme(legend.position = c(0.8, 0.8),
                   strip.text.x = element_text(margin = margin(0.02,0,0.02,0, "cm")))+
             scale_color_manual(name = "Tissue", values = c("steelblue", "orange")))

g_nrads
```


```{r}
ggsave(filename = "Fig_1.pdf",
       plot = g_nrads,
       device = "pdf",
       width = 4,
       height = 3.5)
```

